---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
Load the packages required for the analysis. Load the data by reading the csv file which should be located in the working directory. Convert the date strings into POSIXct dates. Need POSIXct instead of the default result of strptime (POSIXlt) so that can group_by the date in subsequent steps.
```{r echo=TRUE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(pander)
act <- read.csv("./activity.csv")
act$date <- as.POSIXct(strptime(act$date, "%Y-%m-%d"))
```

## What is mean total number of steps taken per day?
Use dplyr and ggplot2 packages to group and summarize data as well as plot the histogram. Remove NAs from the dataset.
```{r echo=TRUE}
dat <- act[complete.cases(act),]
dat <- dat %>% group_by(date) %>% summarise(sum = sum(steps), mean = mean(steps), median = median(steps))

ggplot(dat, aes(x=date, y=sum)) + geom_bar(stat = "identity") + 
  labs(title = "Total Number of Steps Taken Each Day", x = "Date", y = "Number of Steps")

pander(dat)

```


## What is the average daily activity pattern?

```{r echo=TRUE}

```

## Imputing missing values

```{r echo=TRUE}

```

## Are there differences in activity patterns between weekdays and weekends?
